Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-01 manuscript.tex
latest

Some abstract stuff that runs over multiple lines so that I can see the quote formatting.

@@

In section

@@

We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model). All territories are identical, and the total number of territories is $K$. Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity. In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.

@@

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $K\rightarrow \infty$, with $n_i/K$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$. Although $K$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). This dispersal Poisson distribution is denoted $p_i(x_i)$. Note that the large $n_i$, large $K$ approximation places no restrictions on our densities $n_i/K$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).

@@

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $K$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

@@

The ideal environmental archetype is characterized by the near-absence of stress and disturbance. Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$. Moreover, since $L=b_i N/(N-K)$ by definition, population density is also high $N/K\approx 1$. Thus, territorial contests are decisively important.

The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction. Disturbances do not only affect adults as in Eq. \eqref{eq:delttot}, but also juveniles in the process of territorial contest. These juvenile deaths can be represented as a fractional reduction in the number of territories secured. To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration). The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/K\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

@@

The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space. In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense. Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood. In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $K$, not relative to ideal conditions).

@@

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} The realization of Grime's triangle in our model. Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance. The low-$K$ interpretation of stress is also shown. The vertices of the triangles correspond to environmental archetypes. Selection favors different traits at each vertex, leading to different trait clusters.} 
\end{figure}

@@

In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$K$ interpretation of the stress archetype. Lowering $d$ is obviously essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals). For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator. This strategy is associated with resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods. This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$K$ interpretation).

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Yet Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is contrary to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability. The latter form of selection has been termed $\alpha$-selection after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. Unlike the correlation

In contrast to the problems with a strict $r$/$K$ dichotomy, there are

In reality there may be trade-offs between $b$ and $c$ e.g. due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}.

Actual K selection,

Lotka Volterra with alpha,

resource comp.

significance of stage structure

@@

More broadly, while our model can be applied to questions at a species level (e.g. ecological invasions), our focus here is on the evolution of genotype frequencies within a population. This is only possible because the model accounts for the growth of mutants from low densities. Given this focus, our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ that will be sustained in the population. Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases. However, more care would be needed if we were looking species interactions instead of genotype evolution.

eco-evo stuff

@@

For sufficiently large $K$, holding $n_i/K$ fixed, the Poisson limit theorem implies that the binomial distributions for $x_i$ at each successive stage of this procedure are all closely approximated by a Poisson distribution with mean $l_i$, where we have used the fact that large $K$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/K$.

Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$. Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $100$ or less).

>>>>

The classic lottery model is generalized to accomodate arbitrary densities. We use the model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal theory of plant primary strategies. We evaluate the density-dependent lottery model as a general-purpose model of density-dependent selection in the spirit of MacArthur's $r$/$K$ theory.

@@

In section ``Model'' we introduce the basic assumptions of our generalized lottery model. Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices. The following two sections discuss the behaviour of rare mutants and our treatment of Grime's triangle.

@@

We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model). All territories are identical, and the total number of territories is $T$. Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity. In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.

@@

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $K\rightarrow \infty$, with $n_i/K$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$. Although $T$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). This dispersal Poisson distribution is denoted $p_i(x_i)$. Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).

@@

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

@@

The ideal environmental archetype is characterized by the near-absence of stress and disturbance. Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$. Moreover, since $L=b_i N/(N-T)$ by definition, population density is also high $N/T\approx 1$. Thus, territorial contests are decisively important.

The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction. Disturbances do not only affect adults as in Eq. \eqref{eq:delttot}, but also juveniles in the process of territorial contest. These juvenile deaths can be represented as a fractional reduction in the number of territories secured. To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration). The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/T\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

@@

The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $T$ for a given amount of space. In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense. Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood. In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $T$, not relative to ideal conditions).

@@

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (T) \\ \hline
  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
  Density $N/T$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$T$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} The realization of Grime's triangle in our model. Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance. The low-$T$ interpretation of stress is also shown. The vertices of the triangles correspond to environmental archetypes. Selection favors different traits at each vertex, leading to different trait clusters.} 
\end{figure}

@@

In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$T$ interpretation of the stress archetype. Lowering $d$ is obviously essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals). For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator. This strategy is associated with resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods. This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$T$ interpretation).

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability. To avoid this ambiguity, the latter form of selection has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. Unlike saturation density, there is support for a negative relationship between competitive success at high densities and maximal growth rate \cite{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate. However, competitive success as measured by $\alpha$ (i.e. the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structure and territoriality, it also includes the ability of each genotype to produce contestants i.e. $b$ in our model. In contrast, our $c$ is strictly competitive ability only --- as such, it does not directly affect population density (section ``Model'').

$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1962}, would be represented in our model by smaller individual territorial requirements. To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density $L$. The situation is more complicated if those genotypes differ in multiple traits, and when the differences in territorial requirements become large enough that territorial contests can occur on different scales. We leave these complications for future work.

@@

More broadly, while our model can be applied to questions at a species level (e.g. ecological invasions), our focus here is on the evolution of genotype frequencies within a population. This is only possible because the model accounts for the growth of mutants from low densities. Given this focus, our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ that will be sustained in the population. Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases. However, this restriction could become important when looking at species interactions rather than genotype evolution.

In our view, the density dependent lottery model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood). In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that resources are spatially localized (e.g. due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle a competitive ability $c$ should be derivable from resource competition. The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It is possible that explicit resource tracking is in fact not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species. We are not aware of any attempts to delineate the conditions under which explicit resource tracking is not required, let alone to connect those results to the density-dependent selection literature [need to check Tilman's book].

On the other hand, the model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$. The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, without arbitrarily assigning an absolute fitness cost to that competition. [citations to sexual selection stuff?]

@@

For sufficiently large $K$, holding $n_i/T$ fixed, the Poisson limit theorem implies that the binomial distributions for $x_i$ at each successive stage of this procedure are all closely approximated by a Poisson distribution with mean $l_i$, where we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.

Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$. Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $10^2$ or less).
